---
title: "LING 3300 - Week 9: Regression with Categorical Predictors"
author: "Thomas Kettig"
date: "March 12, 2026"
output: html_document
---

contrasts(COG_means$poa) <- contr.sum(2)
contrasts(COG_means$voicing) <- contr.sum(2)

# Check the contrasts to make sure
contrasts(COG_means$poa)
contrasts(COG_means$voicing)

# Making model
fit.sum <- lm(meanCOG ~ voicing * poa, COG_means)
summary(fit.sum)
```
Now let's figure out what those numbers mean.

The intercept here is 6395, which is the **mean of the means of the voicing conditions and the poa conditions**. In this case, because our data is nicely balanced and we have the same number of observations in each of our subsets, the mean of the voiced and voiceless means will be the same as the mean of the alveolar and interdental means: both of these are 6395, as illustrated below:

```{r message = F}

# We'll make a dataset that's just the voiceless means and another that's just the voiced means
just_voiceless <- COG_means %>%
  filter(voicing == "voiceless")
just_voiced <- COG_means %>%
  filter(voicing == "voiced")

# We'll make a dataset that's just the alveolar means and another that's just the interdental means
just_alveolar <- COG_means %>%
  filter(poa == "alveolar")
just_interdental <- COG_means %>%
  filter(poa == "interdental")

# What are the means of each of these subsets?
mean(just_voiceless$meanCOG) # 6957
mean(just_voiced$meanCOG) # 5833
mean(just_alveolar$meanCOG) # 6817
mean(just_interdental$meanCOG) # 5973

# What's the mean of the voiceless and voiced means?
(mean(just_voiceless$meanCOG) + mean(just_voiced$meanCOG)) / 2 # 6395

# What's the mean of the alveolar and voiceless means?
(mean(just_alveolar$meanCOG) + mean(just_interdental$meanCOG)) / 2 # 6395
```
Alright, so that's the intercept. The next line says 'voicing1'. Remember how we've set voiceless to -1 and voiced to 1 in this sum coding? It's using '1' to mean "voiced". So if we take 6395 - 562, we get 5833. Look just above and you'll see that this is the mean of the voiced subset of the data, which includes the alveolar and interdental voiced fricatives. The p-value is telling us that this mean of voiced fricatives is significantly different from the mean of means, meaning that voicing has a significant "main effect" on COG.

The next line says 'poa1'. Remember how we've set alveolar to -1 and interdental to 1 in this sum coding? It's using '1' to mean "interdental". So if we take 6395 - 422, we get 5973. Look just above and you'll see that this is the mean of the interdental subset of the data, which includes the voiceless and voiced interdental fricatives. The p-value is telling us that this mean of interdental fricatives is significantly different from the mean of means, meaning that poa has a significant "main effect" on COG.

Now for something a bit more complicated: The next line says 'voicing1:poa1'. Since '1' in voicing is "voiced" and '1' in poa is "interdental", what this telling us is what the *added effect of being both voiced and interdental is, on top of the main effects of being voiced and being interdental*. Here's what that means:

First, we account for the main effect of voicing: We start with 6395 and add -562 (also known as subtracting 562) to get 5833. 

Then, we need to account for the main effect of poa: We take 5833 and add -422 (also known as subtracting 422), which gives us 5411. 

Last, we'll add in that interaction of -518: 5411 - 518 = 4893. If you look again above, we found that the mean of the voiced interdental was 4894! (The difference of 1 is because I rounded up/down to the nearest Hz in all those calculations, but if you tried the same thing with a larger number of decimals you'd see these results are equal.)

### Emmeans 

In case you're interested in seeing how emmeans does pairwise comparisons all at once, here's a bit of code. We don't have time to get into this, but I encourage you to Google how estimated marginal means work! The numbers sometimes come out a bit different from what the treatment coding model might give you, but because of whatever mysterious math stuff it does. You can find further details online.

```{r message = F}
library(emmeans) # You might need to install this package if you don't already have it
emmeans(fit.dummy, specs = pairwise ~ voicing * poa)

# Should give same results for the sum coded and treatment coded models, just in a different order because our factors were ordered differently in the two models
emmeans(fit.sum, specs = pairwise ~ voicing * poa) 
```

### Practice with interactions

1. Consider this fricatives dataset. We've seen before that there's a generally positive correlation between MaxPF and their M1 (a.k.a. COG): the higher your MaxPF, the higher your M1. 

