---
title: "LING 3300 - Tutorial 5"
author: "Thomas Kettig"
date: "October 1, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, as always, let's load in the packages and dataset we'll be working with for these examples. To start, let's use the basic jury dataset (not the Qualtrics one from last tutorial).

```{r eval = T}
library(tidyverse)
jury <- read.csv("/Users/Thomas/Library/CloudStorage/OneDrive-YorkUniversity/LING 3300/Datasets/Jury/jury_data_basic.csv")
```

### Calculating Standard Error

We've seen before in Tutorial 2 how to get a dataframe with means, standard deviations, and counts of observations:

```{r eval = T}
same_means <- jury %>% 
  group_by(Level, CorrectAnswer) %>% 
  summarize(mean_sameness = mean(SameAnswer), 
            sd_sameness = sd(SameAnswer),
            count = length(SameAnswer))
```

We've also seen in our class lecture that the standard error is calculated by taking the standard deviation and dividing by the square root of the number of observations. That means that once we have a dataframe saved with the means, standard deviations, and counts, all we need to do is add a column that does the standard error calculation:

```{r eval = T}
same_means$se <- same_means$sd_sameness / sqrt(same_means$count)
```

(Let's put aside for the moment that standard error should be calculated based on a set of means of independent sampling distributions, such as within-speaker means, not raw observations. In the practice below, you'll be able to do it the proper way and first calculate sample means, from which you then get the means-of-means and sd-of-means and count-of-means, and then can calculate the standard error.)


### Calculating margin of error + confidence interval

I recommend having your slides from Week 4 open while you read the following.

Let's first read in our L2 lexical decision data and run the code that got us our nice reaction time means-of-means + sd + count + se dataframe.

```{r eval = T}
# Loading in the dataset
lex <- read.csv("/Users/Thomas/Library/CloudStorage/OneDrive-YorkUniversity/LING 3300/Datasets/L2_English_Lexical_Decision_Data.csv")

# Making a character-based column "gender" out of the previous numerical "sex" column
lex$gender <- ifelse(lex$sex == 1, "m", "f")

# Making a column for a (fairly arbitrary) subset of Germanic vs. other dominant languages
lex$Germanic <- ifelse(lex$domLang == "English" |
                         lex$domLang == "German" |
                         lex$domLang == "Norwegian" |
                         lex$domLang == "Danish" |
                         lex$domLang == "Afrikaans",
                       "Germanic", "Other")

# Making a dataframe called RTmeans that has some descriptive stats
RTmeans <- lex %>%  # We start with the overall data
  group_by(Germanic, gender, workerID) %>%  # We group by participant and keep in Germanic and gender
  summarize(meanRT = mean(RT), # We summarize the mean of the reaction times in a new meanRT column
            medianRT = median(RT),  # Plus a column for the median RT
            sdRT = sd(RT),  # And a column for the standard deviation of the RTs
            count = length(RT))  # And a count of how many observations there were

# Now, from that by-speaker list of descriptive stats, we make a dataframe called RTgrandmeans
RTgrandmeans <- RTmeans %>% 
  group_by(Germanic, gender) %>% # We group by Germanic and gender...
  summarize(meanofmeansRT = mean(meanRT), # ...and get a mean of the by-participant means...
            sdofmeansRT = sd(meanRT), # ...and the standard deviation of those by-participant means...
            count = length(meanRT)) # ...and the number of participants in each category

# And also then calculate the standard error of the mean by dividing the std of means by the square root of the counts
RTgrandmeans$se <- RTgrandmeans$sdofmeansRT / sqrt(RTgrandmeans$count)
```

Now that we have the standard error of the means, we can calculate the confidence interval around each of the Germanic/non-Germanic and M/F combinations of reaction time means-of-means. Recall that the confidence interval is the mean ± the margin of error. The formula for margin of error is: standard error times the t-value or z-value for a particular confidence level. Using the t-value rather than the z-value is recommended when there's a relatively low number of samples (under 30 or so), or where we don't know the population standard deviation. Here we have plenty of samples, but as with most measurements in linguistics, we can't really claim to know what the population-wide standard deviation is, so let's stick with t-values.

How do we get a t-value? Well, the R function qt() will give that to us! qt() requires two arguments (see slide 21 from Unit 4 lecture): the point in the t-distribution you want to find the value for and the degrees of freedom in your sample.

Let's say that we want to report a 95% confidence interval, as is most typical in linguistics. Recall that since the difference between 95% and 100% is 5%, that means our alpha is 0.05. Remember how we talked about that 0.05 being split between the upper and lower tails of the normal distribution? That means that some important numbers to keep in mind as we calculate our t-values will be 0.025 and 0.975.

In addition to those numbers, we also need to know our degrees of freedom. To get degrees of freedom, we just take our sample size minus one. So first, let's make a new column that takes our sample size counts and subtracts one, to give us degrees of freedom. You can probably already see that this makes a relatively small difference for our data, where we have hundreds of samples; this subtraction of one would make a bigger impact if we had smaller sample sizes

```{r eval = T}
RTgrandmeans$df <- RTgrandmeans$count - 1
```

Now that we have a *df* column for each of the conditions, we can make another column calculating the lower t-values and the upper t-values.

```{r eval = T}
RTgrandmeans$tval_lower <- qt(0.025, RTgrandmeans$df)
RTgrandmeans$tval_upper <- qt(0.975, RTgrandmeans$df)
```

Now that we have our degrees of freedom and our t-values in columns, we can calculate the margin of error for each of the rows by multiplying the standard error by the t-values.

```{r eval = T}
RTgrandmeans$margin_of_error_lower <- RTgrandmeans$se * RTgrandmeans$tval_lower
RTgrandmeans$margin_of_error_upper <- RTgrandmeans$se * RTgrandmeans$tval_upper
```

Look in the dataframe and see that now you've got the margin of error around the mean that will comprise your confidence interval: plus or minus that particular number. So our last step will be to add or subtract that value from the mean to find the lower and upper bounds of the CI. Since the lower margin of error is already negative, we can *add* that to our mean and it will count as subtraction.

```{r eval = T}
RTgrandmeans$lower95CI <- RTgrandmeans$meanofmeansRT + RTgrandmeans$margin_of_error_lower
RTgrandmeans$upper95CI <- RTgrandmeans$meanofmeansRT + RTgrandmeans$margin_of_error_upper
```

### Visualizing confidence intervals

We have everything we need here to make plots of these confidence intervals. Let's make a plot with the means plotted as points, which we can do with geom_point(). We can add bars for the confidence intervals with geom_errorbar(). We can split the data by Germanic vs. non-Germanic on the x-axis, have the y-axis be reaction time, and separate the genders by colour. For the error bar, we can set ymin (the bar's lowest point) to be the lower 95% CI and the ymax (the bar's highest point) to be the upper 95% CI.

```{r eval = T}
ggplot(RTgrandmeans) +
  geom_point(aes(x = Germanic, y = meanofmeansRT, color = gender)) +
  geom_errorbar(aes(x = Germanic, y = meanofmeansRT, color = gender, ymin = lower95CI, ymax = upper95CI))
```

So far we've been making ggplots by putting the dataframe name inside ggplot() and all the aesthetics inside the individual plotting functions. But do you see how we have the identical aesthetics aes(x = Germanic, y = meanofmeansRT, color = gender) in both of the plotting functions here? When multiple plotting functions take the same dataframe and aesthetics, we can (sometimes) save some space by sticking the parts of aes() that should be applied to all of the functions into the ggplot() function. The following code therefore accomplishes the same thing as above. Like I've been saying, there are usually many different ways to accomplish the same thing in R!
 
```{r eval = T}
ggplot(RTgrandmeans, aes(x = Germanic, y = meanofmeansRT, color = gender)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower95CI, ymax = upper95CI))
```

Anyway -- this plot isn't that great yet, for one thing because our mean points and error bars are overlapping. We use the command *position = position_dodge(width = 0.5)* in both the geom_point() and geom_errorbar() functions in order to spread them apart a bit horizontally. (Unfortunately, it doesn't seem like you can easily stick this command inside the overall ggplot() function even though it's identical for both plotting functions. Sometimes R is just odd that way.)

```{r eval = T}
ggplot(RTgrandmeans, aes(x = Germanic, y = meanofmeansRT, color = gender)) +
  geom_point(position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(ymin = lower95CI, ymax = upper95CI), position = position_dodge(width = 0.5))
```

You can also play around with the size of the mean point, the widths of the error bars, and the width of the error bar lines.

```{r eval = T}
ggplot(RTgrandmeans, aes(x = Germanic, y = meanofmeansRT, color = gender)) +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  geom_errorbar(aes(ymin = lower95CI, ymax = upper95CI), width = 0.15, position = position_dodge(width = 0.4), linewidth = 0.6)
```

Alright. Now our lines of code are getting awfully long! Luckily, we can add some line returns in such a way that the whole thing will run the same way as before! If you have begun a function with an open parenthesis and haven't closed it yet, you can you add as many line breaks as you'd like and R will ignore them and keep reading to figure out what comes next until it gets to the final closing parenthesis. Here's the same code, but spaced out a bit more to make things more legible:

```{r eval = T}
ggplot(RTgrandmeans, 
       aes(x = Germanic,
           y = meanofmeansRT, 
           color = gender)) +
  geom_point(position = position_dodge(width = 0.4), 
             size = 3) +
  geom_errorbar(aes(ymin = lower95CI, ymax = upper95CI),
                width = 0.15,
                position = position_dodge(width = 0.4),
                linewidth = 0.6)
```

It's also rather common to see line returns after some opening parentheses and before some ending parentheses. Here's another way to write the code which might help you keep track of all the parentheses going on, since as soon as you start having lots of line breaks and embedded functions things can get messy.

```{r eval = T}
ggplot(
  RTgrandmeans, 
  aes(
    x = Germanic, 
    y = meanofmeansRT, 
    color = gender
    )
  ) +
  geom_point(
    position = position_dodge(width = 0.4), 
    size = 3
    ) +
  geom_errorbar(
    aes(ymin = lower95CI, ymax = upper95CI), 
    width = 0.15, 
    position = position_dodge(width = 0.4), 
    linewidth = 0.6
    )
```

The number and placement of your line breaks is really a matter of style and comfort -- as long as the code runs correctly without errors, it's up to you how to code, though I do encourage you to break up particularly long lines of code so that they are more readable.

We can also add nicer labels and a title, or beautify the plot with themes:

```{r eval = T}
ggplot(RTgrandmeans, aes(x = Germanic, y = meanofmeansRT, color = gender)) +
  geom_point(position = position_dodge(width = 0.4), size = 3) +
  geom_errorbar(
    aes(ymin = lower95CI, ymax = upper95CI),
    width = 0.15,
    position = position_dodge(width = 0.4),
    linewidth = 0.6
  ) +
  labs(
    x = "Language family",
    y = "Mean RT",
    color = "Gender",
    title = "Mean of means of RT with 95% CI whiskers"
  ) +
  theme_minimal()
```

#### Practice 

In the Qualtrics jury dataset, each participant only gave us one estimate for how familiar they'd rate themselves with regard to each accent (Middlesbrough, Newcastle, or SSBE). Even though we've mostly been discussing the standard error when talking about calculating the mean of means, we don't actually need to find multiple means (such as lots of participant-internal means) to calculate standard error and therefore confidence intervals. 
So let's try it on this Qualtrics dataset: Begin your code the same way you did for the exercise in Tutorial 4, wrangling the data into a format that gives you a column for Accent and a column for Familiarity rating. Now, calculate the mean familiarity rating for each accent, along with the 95% confidence interval around each accent mean. Then make a plot that visualizes these means and CIs. Note that here you won't have to worry about defining colours or dodging the positions -- since we just care about one independent variable here (Accent), this will be a simpler ggplot to make than the above ones where we wanted to show the interaction of two independent variables (Germanic and gender).


<font size="1.5"> Disclaimer: Some of these original materials were put together by Eleanor Chodroff and Elisa Passoni for the University of York. Thomas Kettig then inherited it and modified as needed, particularly based on notes by Nathan Sanders from the University of Toronto. The R software and the packages are distributed under the terms of the GNU General Public License, either Version 2, June 1991 or Version 3, June 2007 (run the command licence () for more information)</font>
